---
## I have commented out the default YAML title and removed the author and data 
## info as I don't really as I don't want them to appear on the title page

# title: "Generic Manuscript Template"

# This sets up the format to match most journal requirements of 1" margins
# and 12pt Times new Roman font. The catch is that to specify the font
# you have to use the xelatex latex_engine. Also important to note, most 
# journals need Word files (or Latex) for copy editing. You can simply
# comment out lines (20, 21, 23 and 24) and uncomment the 
# word_document output. 
# 
geometry: margin=1in
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
    - \usepackage{fontspec}         # can only be used with xelatex output
    - \setmainfont{Times New Roman} # can only be used with xelatex output

output:
  pdf_document:
    latex_engine: xelatex
## Other output options are:
# output: pdf_document
# output: html_document
# output: word_document

## Specify your csl here. You can use local or remote 
## sources of csl files but local is better.
## below are two good sources of CSL files
## https://github.com/citation-style-language/styles
## https://www.zotero.org/styles

# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology-letters.csl
csl: https://www.zotero.org/styles/ecology-letters
# csl: "ecology-letters.csl"

# The .bib file allows you to add citations without Zotero and to share your
# library with co-authors so that everyone has the same list of potential
# references. Co-authors can just add new entries to the .bib file if they think
# a citation is relevant
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set
# I like to put all of my library calls here, each on a seperate line. I am not 
# sure if this is good or bad practice.

```

```{r, include=FALSE,message=FALSE, echo=FALSE}
RequiredPackages <- c("tidyverse","palmerpenguins","GGally","kableExtra","medicaldata","rmarkdown",
                      "dataverse","tibble","dplyr","ggplot2","mgcv","cowplot")
for (i in RequiredPackages) { #Installs packages if not yet installed
  if (!require(i, character.only = TRUE)) install.packages(i)
}

library("tidyverse")
library("palmerpenguins")
library("GGally")
library("kableExtra")
library("medicaldata")
library("rmarkdown")
library("dataverse") 
library("tibble") 
library("dplyr")
library("ggplot2") 
library("mgcv") 
library("cowplot") 

knitr::opts_chunk$set(echo = FALSE)
```

**REPRODUCIBILITY MINI PROJECT: ALBERTA TREES**

Rolando Trejo Pérez 1^\*

1.  Institut de recherche en biologie végétale, Université de Montréal

\*Corresponding Author

email: rolando.trejo.perez@umontreal.ca

**Author Contributions:** RTP is writting this mini reproducibility project.

**Data Availability:**Two databases were downloaded from the Seasonal and annual dynamics of western Canadian boreal forest plant communities: a legacy dataset spanning four decades:https://borealisdata.ca/dataset.xhtml?persistentId=doi:10.5683/SP3/PZCAVE


**Conflict of Interest statement**

No conflicts of interest

**Acknowledgements:** I thank George H. La Roi and colleagues, Amelia Hesketh, Jenna Loesberg,Ellen Bledsoe, Justine Karst,and Ellen Macdonald.

\newpage

# Abstract

The goal of this manuscript is using reproducibility workflow in ecology and evolution. I used open access data, which is available in Boreales (see references below). This data was rescued by Hesketh et al., 2021. This mini project was created with the open-source software R, OSF and GitHub. A very simple question is stated: does species richness varies along the soil temperature? All the code used here is available in a public GitHub repository.

**Key-words**: reproducibility, diversity, species richness, soil temperature.
\newpage

# Introduction

Global warming is predicted to significantly alter species physiology, biotic interactions and thus ecosystem functioning, as a consequence of coexisting species exhibiting a wide range of thermal sensitivities. The richness diversity of plants decreased with increasing soil temperature, driven by decreasing plant species richness (Robinson et al., 2018). Species richness and its relationship with temperature has been assessed along altitudinal and horizontal gradients revealing significant linear relationships between species richness and altitude and climatic variables (Pickering et al., 2008). 


Here, in this manuscript, I used a very simple linear model to explore the relationship between vascular diversity and soil temperature from  the Seasonal Dynamics (SEADYN) and later Annual Dynamics (ANNDYN) research project. The primary purpose of this project was to document seasonal changes in the vegetative composition during the snow-free season (May through October) and longer-term changes in vegetation and forest mensuration for boreal forest stands in Alberta, Canada dominated by Pinus banksiana (Lamb.) (Hesketh et al., 2021).

# Methods

This mini reproducibility project was built using rescued data by Amelia Hesketh, Jenna Loesberg, Ellen Bledsoe, Justine Karst,and Ellen Macdonald in 2021 from an Alberta legacy dataset spanning four decades (1980-2015). We use two different datasets: 1) Hondo Vascular Cover (1980-2015) and 2) Hondo Soil Temperature (1980-2010). These datasets are available in Borealis at https://borealisdata.ca/dataset.xhtml?persistentId=doi:10.5683/SP3/PZCAVE

A very simple model is explored in this reproducible project: Does species richness varies along the soil temperature? For this purpose i only use 2010 data and the following linear model approach:

$$
SR = ST+e
$$

where $SR$ denotes the species richness, $ST$ is the soil temperature in Celsius,and  $e$ is the error.

This mini project was created with the open-source software R. Packages used were dataverse,tibble,dplyr,ggplot2, and mgcv.
Statistical analyses were carried out in R 3.5.0 (R Core Team 2017). All code used in this manuscript is available on GitHub <https://github.com/CIEE-Living-Data-Project/Rolando_Trejo_Reproducibility_LDP_2022>.

# Results

```{r code, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
### 1. CREATING DIRECTORIES AND INSTALLING R PACKAGES ###

RequiredPackages <- c("dataverse","tibble","dplyr","ggplot2","mgcv")
for (i in RequiredPackages) { #Installs packages if not yet installed
  if (!require(i, character.only = TRUE)) install.packages(i)}

library("dataverse") # To access to data from dataverse repository
library("tibble") # To see data frames in tidyverse-form
library("dplyr") # To solve the most common data manipulation challenges
library("ggplot2") # To data visualization
library("mgcv") # To estimate penalized Generalized Linear models including 
                #Generalized Additive Models and Generalized Additive Mixed Models

# 1.1 Check for directory path

getwd()

### 2. DOWNLOAD DATA FROM DATAVERSE (download in 2022-15-09): ###

# Two databases were download from the Seasonal and annual dynamics of western 
# Canadian boreal forest plant communities: a legacy dataset spanning four decades:

# [https://borealisdata.ca/dataset.xhtml?persistentId=doi:10.5683/SP3/PZCAVE]

# 2.1 Vascular cover from 1980 to 2015

Hondo_VascularCover_1980_2015 <- get_dataframe_by_name(
  filename = "Hondo_VascularCover_1980_2015.tab",
  dataset = "10.5683/SP3/PZCAVE", 
  server = "https://borealisdata.ca/dataverse/ubc")

# 2.2 Soil temperature from 1980 to 2010 

Hondo_SoilTemp_1980_2010 <- get_dataframe_by_name(
  filename = "Hondo_SoilTemp_1980_2010.tab",
  dataset = "10.5683/SP3/PZCAVE", 
  server = "https://borealisdata.ca/dataverse/ubc")
### 4. DATA MANIPULATION ###
    
# 4.1. Generate a subset of data considering only 2010 data to simplify the statistical analyses.
    
    Hondo_VascularCover_2010 <- subset(Hondo_VascularCover_1980_2015,
                                  year== "2010" )          # Selecting from one category in rows
    Hondo_SoilTemp_2010 <- subset(Hondo_SoilTemp_1980_2010,
                                  year== "2010" )
# 4.2. Merge Vascular cover and Soil temperature datasets by month, year, stand and quad 
   
    df=inner_join(Hondo_VascularCover_2010, Hondo_SoilTemp_2010, by = c("month","year","stand", "quad"))

# 4.3 It creates a subset ob non-binary data

df_non_binary <- subset(df,select = c(stand, month,year,date,quad,temp_C))

#  It creates a subset of binary data: abundances needs to be transformed into 0 and 1 
# in order to count the species richness in each row.

df_species <- subset(df,select = c(-stand, -month,-year,-date,-temp_C))

# It transforms abundances into 0 and 1

df_binary <- data.frame(df_species[1], (df_species[-1] > 0) * 1)

# It merges the non binary data, soil temperature, and species abundances transformed into 0 and 1.

df_trans <- data.frame(df_non_binary,df_binary)

# It sums the number of species in each row

df_trans$species_no <- rowSums( df_trans[,8:221] )

# It create the final table version: it includes only the species richness and soil temperature

Species_Temp <- subset(df_trans,select = c(species_no,temp_C))

# Linear model or not?

library(mgcv)

# The following code provides a linear and non-linear model approach:

linear_model <- gam(species_no  ~ temp_C, data = Species_Temp)
smooth_model <- gam(species_no  ~ s(temp_C), data = Species_Temp)
# Visualization

library(ggplot2)

# It graphs the linear model (species richness vs soil temperature) using ggplot2

Linear<-ggplot(Species_Temp, aes(x = temp_C, y = species_no )) +
  geom_point() +
  geom_line(colour = "red", size = 1.2,
            aes(y = fitted(linear_model))) +
  xlab("Soil temperature (C)") +
  ylab("Species richness") +
  labs(title = "A) Linear model line")+
  theme_bw()

Linear

# It graphs the non-linear model (species richness vs soil temperature) using ggplot2

Non_linear<-ggplot(Species_Temp, aes(x = temp_C, y = species_no )) +
  geom_point() +
  xlab("Soil temperature (C)") +
  ylab("Species richness") +
  labs(title = "B) Smooth model curve")+
  geom_line(colour = "blue", size = 1.2,
            aes(y = fitted(smooth_model))) +
  theme_bw()

Non_linear
```
A higher species richness is linked to a lower soil temperature according to this simple linear model. We can also see that linearity is supported by these data (see Figure 1). However, this is just a model considering soil temperature as the only predictor. Other variables must be considered to search if it is a real observed trend. See more detail in the [Living Data Tutorials](https://living-data-tutorials.github.io/website/)

# Discussion

The code used here in this mini reproducibility project can be used as guide through the creation and management of a fully reproducible manuscript using RMarkdown and Rstudio.

# References

```{=html}
<!-- this littel code block is a comment that won't show up in the knitted 
document. It is a good way to add hidden things and comments to your RMarkdown 
document. For example... -->
```
```{=html}
<!-- this little chunk of code (the ::: stuff below) forces the references to show up here. If you
don't use this then they are put at the end of the document by default -->
```
Hesketh, A., Loesberg, J., Bledsoe, E., Karst, J., & Macdonald, E. (2021). Seasonal and annual dynamics of western Canadian boreal forest plant communities: A legacy dataset spanning four decades [Data set]. Scholars Portal Dataverse. https://doi.org/10.5683/SP3/PZCAVE

Pickering, C., Hill, W. & Green, K. Vascular plant diversity and climate change in the alpine zone of the Snowy Mountains, Australia. Biodivers Conserv 17, 1627–1644 (2008). https://doi.org/10.1007/s10531-008-9371-y

Robinson, S.I., McLaughlin, O.B., Maereinsdóttir, B., and O'Gorman, E.J. Soil temperature effects on the structure and diversity of plant and invertebrate communities in a natural warming experiment, Journal of Animal Ecology, 87(3):634-646 (2018).
\newpage
```{r figure_02, echo=FALSE, message=FALSE, warning=FALSE}

library(cowplot)
plot_grid(Linear,Non_linear, labels = "") #"AUTO"

```

Figure 1. A) linear model approach and B) a non linear model approach. Linearity is respected.

\newpage

# Tables

There are not table to show.

\newpage

# Appendices

There are not appendices to show.
